---
title: "Insect Body Mass & Size Database — Assembly and Exploration"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5
)
```

# 1. Introduction

This document describes the assembly, cleaning, and initial exploration of a
multi-source insect body mass and size database. Data are drawn from a
configuration file (`dataset.csv`) that specifies the path, format, and
metadata for each contributing dataset. All records are standardised into a
common schema before taxonomic annotation via ncbi and filtering to Insecta.

---

# 2. Setup

```{r packages}
library(readxl)
library(tidyverse)
library(taxizedb)
library(parallel)
```

---

# 3. Data Loading

## 3.1 Configuration

The configuration file drives all downstream loading. Each row describes one
source dataset: its file path, format, relevant columns, trait type, units,
and collection/estimation metadata.

```{r load-config}
config <- read.csv("../data/dataset.csv",
                   stringsAsFactors = FALSE,
                   na.strings       = c("", "NA"),
                   quote            = "\"",
                   strip.white      = TRUE)

cat(sprintf("Configuration file loaded: %d datasets described.\n", nrow(config)))
knitr::kable(
  config %>% select(dataset_name, file_type, trait, metric, state, estimate, doi),
  caption = "First 10 rows of the configuration file"
)
```

## 3.2 Helper Functions

```{r helper-functions}
load_dataset <- function(row) {
  if (row$file_type == "manual") return(NULL)
  cat(paste("Loading:", row$dataset_name, "\n"))

  if (row$file_type == "csv") {
    sep_char <- switch(as.character(row$separator),
                       "comma" = ",", "tab" = "\t", ",")
    data <- if (!is.na(row$encoding))
      read.csv(row$file_path, sep = sep_char, encoding = row$encoding)
    else
      read.csv(row$file_path, sep = sep_char)

  } else if (row$file_type == "xlsx") {
    sheet_num <- ifelse(is.na(row$sheet), 1, as.numeric(row$sheet))
    data <- read_xlsx(row$file_path, sheet = sheet_num)

  } else if (row$file_type == "rdata") {
    env      <- new.env()
    load(row$file_path, envir = env)
    obj_name <- ls(env)[1]
    data     <- get(obj_name, envir = env)
  }

  if (!is.na(row$post_processing)) {
    if (row$post_processing == "calculate_mean_row") {
      columns       <- names(data)[4:224]
      data[data == "NA"] <- NA
      data[columns] <- lapply(data[columns], as.numeric)
      data$mean_row <- rowMeans(
        data[, c("FoL_var_male_average", "FoL_var_female_average",
                 "FoL_HR_average", "FoL_Ten_average")], na.rm = TRUE)
    } else if (row$post_processing == "fix_header") {
      colnames(data) <- data[2, ]
      data           <- data[-c(1:2), ]
    }
  }

  if (!is.na(row$filter_condition))
    data <- data %>% filter(eval(parse(text = row$filter_condition)))

  list(data = data, config = row)
}

check_data <- function(data) {
  data <- data %>%
    filter(complete.cases(data$Species),
           !grepl("[0-9]|(sp|spp)$|(sp\\.|spp\\.)", data$Species),
           Value > 0)
  data$Species <- str_replace_all(data$Species, "\\s*\\([^\\)]+\\)", "")
  data
}

bodymass <- function(file, species, value, trait, metric, state, estimate, doi) {
  state    <- ifelse(is.na(state),    "NA", state)
  estimate <- ifelse(is.na(estimate), "NA", estimate)
  doi      <- ifelse(is.na(doi),      "NA", doi)

  if (!state    %in% c("Dry", "Live", "NA")) stop("Invalid collection type.")
  if (!estimate %in% c("Yes", "No",   "NA")) stop("Invalid estimate type.")
  if (!metric   %in% c("mg", "g", "kg", "cm", "mm")) stop("Invalid metric.")
  if (!trait    %in% c("mass", "size")) stop("Invalid trait.")

  species <- iconv(species, from = "latin1", to = "UTF-8", sub = "")

  tmp <- data.frame(
    Species    = species,
    Trait      = rep(trait,    length(species)),
    Value      = as.numeric(value),
    Metric     = rep(metric,   length(species)),
    Collection = rep(state,    length(species)),
    Estimate   = rep(estimate, length(species)),
    doi        = rep(doi,      length(species)),
    stringsAsFactors = FALSE
  )
  tmp$Species <- sub("_", " ", tmp$Species)
  tmp         <- na.omit(check_data(tmp))
  rbind(file, tmp)
}
```

## 3.3 Ingestion Loop

```{r ingest, results='hide'}
file <- c()

for (i in seq_len(nrow(config))) {
  row <- config[i, ]

  if (row$file_type == "manual") next

  tryCatch({
    info <- load_dataset(row)
    if (is.null(info)) next
    file <- bodymass(file,
                     info$data[[info$config$species_col]],
                     info$data[[info$config$value_col]],
                     info$config$trait,   info$config$metric,
                     info$config$state,   info$config$estimate,
                     info$config$doi)
  }, error = function(e)
    cat(paste("Error loading", row$dataset_name, ":", e$message, "\n")))
}
```

---

# 4. Raw Dataset Overview

```{r raw-summary}
cat(sprintf("Total records      : %d\n",         nrow(file)))
cat(sprintf("Unique species     : %d\n",          n_distinct(file$Species)))
cat(sprintf("Traits represented : %s\n",
            paste(unique(file$Trait), collapse = ", ")))
cat(sprintf("Metrics represented: %s\n",
            paste(unique(file$Metric), collapse = ", ")))
```

```{r raw-table}
knitr::kable(
  file %>%
    group_by(Trait, Metric, Collection, Estimate) %>%
    summarise(n_records = n(), n_species = n_distinct(Species), .groups = "drop") %>%
    arrange(Trait, Metric),
  caption = "Record counts by trait, metric, collection state, and estimate flag"
)
```

## 4.1 Records per Source

```{r records-per-source, fig.cap="Number of records contributed by each source DOI (top 20)."}
file %>%
  count(doi, sort = TRUE) %>%
  slice_head(n = 20) %>%
  mutate(doi = str_trunc(doi, 55)) %>%
  ggplot(aes(x = reorder(doi, n), y = n)) +
  geom_col(fill = "steelblue", alpha = 0.85) +
  coord_flip() +
  labs(x = NULL, y = "Number of records",
       title = "Top 20 contributing sources") +
  theme_minimal(base_size = 11)
```

## 4.2 Value Distributions

```{r value-dist, fig.cap="Log10-transformed value distributions split by trait and metric."}
file %>%
  mutate(log_value = log10(Value)) %>%
  ggplot(aes(x = log_value, fill = Metric)) +
  geom_histogram(bins = 50, alpha = 0.75, colour = "white") +
  facet_wrap(~Trait, scales = "free") +
  labs(x = "log10(Value)", y = "Count",
       title = "Distribution of raw measurements by trait") +
  theme_minimal(base_size = 11)
```

## 4.3 Species Coverage — Paired Measurements

Species with both a measured (non-estimated) size **and** a measured mass are
the most analytically valuable for allometric modelling.

```{r paired-coverage}
species_count <- file %>%
  filter(Estimate %in% c("Yes", "No")) %>%
  group_by(Species) %>%
  filter(
    any(Trait == "size" & Estimate == "No") &
    any(Trait == "mass" & Estimate == "No")
  ) %>%
  distinct(Species) %>%
  nrow()

cat(sprintf(
  "Species with both measured (non-estimated) size AND mass: %d\n",
  species_count))
```

---

# 5. Taxonomic Annotation

Species names are resolved against ncbi via `taxizedb` to retrieve class,
order, suborder, family, genus, and species-level taxonomy. The dataset is
then filtered to **Insecta**.

```{r taxonomy, cache=TRUE}
insect_mass <- file
colnames(insect_mass)[c(1, 7)] <- c("species", "source_doi")

ids      <- name2taxid(unique(insect_mass$species),
                       out_type = "summary", db = "ncbi")
classes  <- classification(ids$id)

process_df <- function(df) {
  df %>%
    t() %>%
    as_tibble() %>%
    slice(-c(2, 3)) %>%
    set_names(make.unique(as.character(df$rank))) %>%
    as.data.frame(stringsAsFactors = FALSE)
}

processed_data <- mclapply(classes, process_df, mc.cores = detectCores())
taxonomy       <- bind_rows(processed_data)

taxonomy    <- taxonomy[, c("class", "order", "suborder",
                             "family", "genus", "species")]
insect_mass <- merge(taxonomy, insect_mass, by = "species")
insect_mass <- insect_mass %>% filter(class == "Insecta")

cat(sprintf("Records after filtering to Insecta: %d\n",    nrow(insect_mass)))
cat(sprintf("Unique insect species             : %d\n",    n_distinct(insect_mass$species)))
cat(sprintf("Unique insect species with mass: %d\n", length(unique(insect_mass$species[insect_mass$Trait=="mass"]))))
cat(sprintf("Unique insect species with measured, not estimated, mass: %d\n", length(unique(insect_mass$species[insect_mass$Estimate=="No" & insect_mass$Trait=="mass"]))))
```

---

# 6. Insect Dataset Exploration

## 6.1 Taxonomic Breakdown

```{r tax-summary-table}
knitr::kable(
  insect_mass %>%
    group_by(order) %>%
    summarise(
      n_records = n(),
      n_species = n_distinct(species),
      n_families = n_distinct(family)
    ) %>%
    arrange(desc(n_species)),
  caption = "Records and species richness per insect order"
)
```

```{r order-barplot, fig.cap="Species richness per insect order (orders with ≥5 species shown)."}
insect_mass %>%
  group_by(order) %>%
  summarise(n_species = n_distinct(species)) %>%
  filter(n_species >= 5) %>%
  ggplot(aes(x = reorder(order, n_species), y = n_species)) +
  geom_col(fill = "forestgreen", alpha = 0.8) +
  coord_flip() +
  labs(x = NULL, y = "Number of species",
       title = "Species richness by insect order (≥5 species)") +
  theme_minimal(base_size = 11)
```

## 6.2 Body Mass Distribution

```{r mass-dist, fig.cap="Log10 body mass distributions by order (mass records only, top 10 orders by record count)."}
top_orders <- insect_mass %>%
  filter(Trait == "mass") %>%
  count(order, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(order)

insect_mass %>%
  filter(Trait == "mass", order %in% top_orders) %>%
  mutate(log_mass = log10(Value)) %>%
  ggplot(aes(x = log_mass, fill = order)) +
  geom_density(alpha = 0.55, colour = NA) +
  facet_wrap(~order, scales = "free_y", ncol = 3) +
  labs(x = "log10(body mass)", y = "Density",
       title = "Body mass distributions across top 10 insect orders") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")
```

## 6.3 Metric Standardisation Overview

```{r metric-summary}
knitr::kable(
  insect_mass %>%
    group_by(Trait, Metric) %>%
    summarise(
      n          = n(),
      mean_value = round(mean(Value, na.rm = TRUE), 4),
      median     = round(median(Value, na.rm = TRUE), 4),
      sd         = round(sd(Value, na.rm = TRUE), 4),
      min        = round(min(Value, na.rm = TRUE), 6),
      max        = round(max(Value, na.rm = TRUE), 2),
      .groups    = "drop"
    ),
  caption = "Summary statistics by trait and metric (pre-unit conversion)"
)
```

## 6.4 Estimated vs. Measured Records

```{r estimate-breakdown, fig.cap="Proportion of estimated vs. measured records per trait."}
insect_mass %>%
  filter(Estimate %in% c("Yes", "No")) %>%
  count(Trait, Estimate) %>%
  group_by(Trait) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = Trait, y = pct, fill = Estimate)) +
  geom_col(position = "stack", alpha = 0.85, colour = "white") +
  scale_fill_manual(values = c("No" = "steelblue", "Yes" = "firebrick")) +
  labs(x = "Trait", y = "Percentage of records (%)",
       fill = "Estimated?",
       title = "Proportion of estimated vs. directly measured records") +
  theme_minimal(base_size = 11)
```

---

# 7. Session Information

```{r session-info}
sessionInfo()
```